---
title: 'Value of Information for Health Economic Evaluations: book code resources'
author: ""
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
 \usepackage[utf8]{inputenc}
 %\VignetteIndexEntry{Value of Information for Health Economic Evaluations: book code resources}
 %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r,echo=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


This is a guide to the code and software resources for the book _Value of Information for Health Economic Evaluations_ (eds. Heath, Jackson, Kunst).  The book is under development, and when published, these resources will be linked from here.

Before running any of this code, you should install the `voi` package, which when the book is published will be available from CRAN. 

```{r}
install.packages("voi")
```

Then you can set the working directory as follows 

```{r}
setwd(system.file("Chemotherapy_Book",package="voi"))
```

then your working directory should contain subdirectories that include `01_data_raw`, `02_data`, etc.  You can check this by running the command `list.files()` in your R console.

Alternatively, you can copy the contents of the `Chemotherapy_Book` directory to your preferred working directory, and work there instead.  If you work in RStudio then it is recommended to use a "project".  You could create a project for this work by going to `File`,`New Project`,`Version Control`,`Git`, then selecting [https://github.com/convoigroup/Chemotherapy_Book](https://github.com/convoigroup/Chemotherapy_Book) as the `Repository URL`, and changing the project name and save location according to your preference.

# Chapter 3: Chemotherapy case study 

The code to run the baseline cost-effectiveness analysis (TODO REF SECTION NUMBERS) is in the `04_analysis/02_baseline_model_output.R` subdirectory ([link to GitHub source](https://raw.githubusercontent.com/convoigroup/Chemotherapy_Book/main/04_analysis/02_baseline_model_output.R)).  This can be run all at once using 

```{r}
source("04_analysis/02_baseline_model_output.R")
```

Or alternatively, open this file in R, and examine and run each line of code by hand. 


# Chapter 4: Value of Perfect Information

#### EVPI analysis for the Chemotherapy model (REF section)

`04_analysis/03_Expected_Value_of_Perfect_Information.R` ([link to GitHub source](https://raw.githubusercontent.com/convoigroup/Chemotherapy_Book/main/04_analysis/03_Expected_Value_of_Perfect_Information.R))

#### EVPPI analysis for the Chemotherapy model  (REF section) 

`04_analysis/04_Expected_Value_of_Partial_Perfect_Information.R` ([link to GitHub source](https://raw.githubusercontent.com/convoigroup/Chemotherapy_Book/main/04_analysis/04_Expected_Value_of_Partial_Perfect_Information.R))


#### Nonparametric regression examples

Code to draw illustrations of GAM regression (REF FIG NOS)
```{r}
misc_dir <- setwd(system.file("book_misc",package="voi"))
```

Code to draw illustrations of Gaussian process regression (REF FIG NO)
```{r}
file.path("misc_dir", "gp_graphs.R")
```

Code to calculate EVPPI using nonparametric regression (REF SECTION NO)
```{r}
file.path("misc_dir", "evppi_reg.Rmd")
```


# Chapter 5: Value of Sample Information


#### Moment matching methods 

`04_analysis/05_Expected_Value_of_Sample_Information_MM.R` ([link to GitHub source](https://raw.githubusercontent.com/convoigroup/Chemotherapy_Book/main/04_analysis/05_Expected_Value_of_Sample_Information_MM.R))


#### Regression-based methods

`04_analysis/05_Expected_Value_of_Sample_Information_RB.R` ([link to GitHub source](https://raw.githubusercontent.com/convoigroup/Chemotherapy_Book/main/04_analysis/05_Expected_Value_of_Sample_Information_RB.R))


#### Presenting EVSI and ENBS Analyses

For base R functions to reproduce the plots in the book, see the [source on GitHub](https://raw.githubusercontent.com/convoigroup/Chemotherapy_Book/main/04_analysis/06_Plotting_Book.R)

For ggplot2 plotting functions, see the [voi package vignette](plots.html) about ENBS and plotting. 